---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Check for and, if necessary, download files in current working directory
unzip file if activity.csv doesn't already exist
```{r, echo = T}
# check if zipfile has been download to current working directory
# if not, download zipfile using method = "curl" for OS X and https
zipfile = "activity.zip"
if(!file.exists(zipfile)) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                      destfile = zipfile,
                      method = "curl")
}

# check if the activity.csv has been unzipped from activity.zip
# unzips activity.zip if not and creates activity.csv in working directory
filename = "activity.csv"
if(!file.exists(filename)) {
        unzip("activity.zip")
}
```

loading csv file into data.frame and preprocess date column for later use
```{r, echo=TRUE}
# load data into variable
active.data <- read.csv("activity.csv")

# cast date column as POSIXct
# allows for extraction of day names later on
active.data$date <- as.POSIXct(active.data$date)
```


## What is mean total number of steps taken per day?

1. Make histogram of the total number of steps taken each day
```{r, echo=T}
# make a 61 levels for each day by 288 for each time entry factor
f <- gl(61, 288)
active.data$dayFactor <- f

# create a vector with the total steps for each of the 61 days
stepsPerDay <- tapply(active.data$steps, active.data$dayFactor, sum)

# plot number of steps per day
barplot(stepsPerDay, xlab = "Day #", ylab = "# of Steps")

## save copy of figure in directory
# check for directory and create if needed
if(!file.exists("./figure")) {
        dir.create("./figure")
}

# add a copy of the plot to the figure directory
dev.copy(png, file = "./figure/AvgSteps-Over-TimePeriod",
         width = 480, height = 480)
dev.off()
```

2. Calculate the median total number of steps taken per day
**Absolute Mean** *total steps recorded over 61 days*
To calculate asbolute mean over 61 day period,
we take the total number of steps recorded and divide by the 61 days
```{r, echo=TRUE}
# number of days in October [31] and November [30]
totalDays = 61

# total number of steps recorded
# set argument na.rm to TRUE to avoid non-numeric values
totalSteps = sum(active.data$steps, na.rm = TRUE)
absMean <- totalSteps / totalDays

# print results
paste("The absolute mean of recorded steps over the entire",totalDays,
      "peroid is", absMean, sep=" ")
```

**Relative Mean** *number of steps taken per day in which data exists*
However,
Considering that there are some missing days,
the mean over days which we have data on would be of greater significance
```{r, echo=TRUE}
# number of days in October [31] and November [30]
totalDays = 61

# count number of missing values
missing <- sum(is.na(active.data$steps))

# account for 288 5-minutes recordings per day
relativeDays <- totalDays - (missing / 288)

# total number of steps recorded
# set argument na.rm to TRUE to avoid non-numeric values
totalSteps = sum(active.data$steps, na.rm = TRUE)
absMean <- totalSteps / relativeDays

# print results
paste("The mean total number of steps taken per day of recorded activity was",
      absMean,"average steps per day during the", relativeDays,
      "days of recorded activity over the 61 days peroid", sep = " ")
```


## What is the average daily activity pattern?

TODO Calc max time period

1. Make a time series plot of the 5-minute interval and the average number
of steps taken, averaged across all days
```{r, echo=TRUE}
# create index column of factors for each of the 288 5-minute peroids in a day
active.data$timeFactor <- c(1:288)

# calculate mean of each peroid, excluding all NA values
means <- tapply(active.data$steps, active.data$timeFactor, mean, rm.na = TRUE)

# plot average amount of steps during each of the 288 5-minute intervals
plot(active.data$interval[1:288], means, type = "l", xlab = "5-minute period",
     ylab = "Average number of Steps (during given period)")

# check for directory and create if needed
if(!file.exists("./figure")) {
        dir.create("./figure")
}

# add a copy of the plot to the figure directory
dev.copy(png, file = "./figure/AvgSteps-Over-TimePeriod",
         width = 480, height = 480)
dev.off()
```

2. Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps?

```{r, echo=T}
# one of two ways to calculate the largest average time period
# is to pull the first index from a reverse ordering
firstIndx <- order(-means)[1]
first <- active.data$interval[firstIndx]

# An addditional way would be to pull the last or tail index from ordered means
secondIndx <- tail(order(means), n = 1)

# both of which give us index 104 which amounts to time period 835
paste("The 5-minute interval with the largest average across all days in the",
      "data set is time period", active.data$interval[secondIndx], sep = " ")
```

## Imputing missing values

My *strategy* for filling in the missing or NA values will be
to use the mean steps for that given time period.
My inferential logic and subsequent assumption being that the number of steps 
one takes will have more variance across different times of the day rather than
the same time across different days.
I attribute this to humans being animals of habit and ritual.
On average, we maintain our sleeping schedule, working schedule and even regular
hobbies and exercise at reoccuring times throughout our days and weeks.
```{r, echo=T}

```


## Are there differences in activity patterns between weekdays and weekends?

TODO update this to be two rows of graphs:
1 with weekdays over time periods
1 with weekends over time periods

```{r, echo=TRUE}
# create index column of factors for each of the 288 5-minute peroids in a day
active.data$timeFactor <- c(1:288)

# calculate mean of each peroid, excluding all NA values
means <- tapply(active.data$steps, active.data$timeFactor, mean, rm.na = TRUE)

# plot average amount of steps during each of the 288 5-minute intervals
plot(active.data$interval[1:288], means, type = "l", xlab = "5-minute period",
     ylab = "Average number of Steps (during given period)")

# check for directory and create if needed
if(!file.exists("./figure")) {
        dir.create("./figure")
}

# add a copy of the plot to the figure directory
dev.copy(png, file = "./figure/AvgSteps-Over-TimePeriod",
         width = 480, height = 480)
dev.off()
```
